Reliable Inferences from Simply Parameterised Models

### 2013-04-06 ms.tex
Fixed typo
Some random writing ont he plane

\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\parindent=0cm
\newcommand{\yy}{\mathbf{y}}

@@

\begin{abstract}
We discuss the practical use of simply parameterised models in Bayesian data analysis, particularly in astronomy.
The use of simply parameterised models is problematic when they are fit to large, informative data sets, resulting in unrealistically small uncertainties on inferences. In common parlance, the formal statistical errors may be small, but the systematic errors are often large and in many cases, difficult to quantify at all. In principle, the best solution to this problem is to develop more realistic complex models that accurately describe our prior beliefs. However, this is often infeasible in practice, due to computational limitations and the difficulty of defining satisfactory prior distributions over high dimensional spaces. As a pragmatic alternative, we show that it is possible to reinterpret the use of simple models in a way that avoids the problem of overconfident
inferences. The key idea is to weaken the connection between the parameters and the data that is assumed in the prior probabilities. This weakening can be implemented straightforwardly using the Nested Sampling algorithm.
\end{abstract}

@@

When the number of parameters being inferred is 3 or greater, enumeration of
all possible values becomes computationally prohibitive. In this situation,
Markov Chain Monte Carlo (MCMC) methods are commonly used to produce samples
drawn from the posterior distribution. These samples can then be used to view
the marginal distribution for any particular parameter(s) of interest.
In addition, if the
samples are viewed as a movie, they can give the viewer an intuitive sense of
the uncertainty in the determination of the parameters: any feature of the
models that varies between movie frames is a conclusion that is uncertain, and any
feature that is common to all frames in the movie is a feature that has a high
posterior probability.

@@

\section{An Example: Fitting an Emission Line}
Throughout this paper we will use a simple example of fitting an emission line
to noisy data to explore the properties of simply parameterised models. The
simulated data was generated from the following complex model:
\begin{eqnarray}
\mathcal{M}_0(x) \propto 3\exp\left[-\frac{1}{2}\left(\frac{x}{0.3}\right)^2\right]
+ \exp\left[-\frac{1}{2}\left(\frac{x - 1}{1}\right)^2\right]
+ 5\exp\left[-\frac{1}{2}\left(\frac{x + 1}{0.5}\right)^2\right]
%M = 3*exp(-0.5*(x/0.3)**2) + exp(-0.5*(x-1.)**2) + 5*exp(-0.5*((x+1)/0.5)**2)
\end{eqnarray}
The fluxes were evaluated at $x$-values ranging from $-10$ to $10$ in steps
of $0.001$, and then normalised to sum to 1:
\begin{eqnarray}
\sum_{i=1}^N \mathcal{M}_0(x_i) &=& 1.
\end{eqnarray}
To simulate noise in the measurement process, errors $\{\epsilon_i\}$ were added, generated
independently from a normal distribution with mean 0 and standard deviation
$\sigma = 3 \times 10^{-4}$.
This produces noisy data $\mathbf{y} = \{y_1, y_2, ..., y_N\}$:
\begin{eqnarray}
\begin{array}{lccr}
y_i = \mathcal{M}_0(x_i) + \epsilon_i, & & & \epsilon_i \sim \mathcal{N}(0, \sigma^2).
\end{array}
\end{eqnarray}
The simulated data is shown in Figure~\ref{emission_line}.

@@

\subsection{The Failure of the Simply Parameterised Model}
We ran the simply parameterised model on the simulated data using the Diffusive
Nested Sampling algorithm \citep{dnest}
\footnote{A C++ implementation of Diffusive Nested Sampling is available
online at {\tt http://github.com/eggplantbren/DNest3/}.}. This provides samples
from the posterior distribution for the parameters $A$, $c$ and $w$. Since these
parameters do not ``exist'' in reality (i.e. the data were generated from the
complex model), it does not make much sense to look at the posterior distribution
for the parameters. However, consider the total flux:
\begin{eqnarray}
F = \sum_{i=1}^N \mathcal{M}(x_i)
\end{eqnarray}
This can be computed both for the true underlying complex model, and the fitted
simple models. In other words, it ``exists''. We computed the posterior distribution
for the total flux using the posterior samples. These are plotted in
Figure~\ref{failure}.

A commonly suggested technique \citep[e.g.][]{gregory} to avoid this situation
is to add an ``error-bar boost'' parameter which we will call $\kappa$.
This adds one more parameter to the list, and the likelihood becomes:
\begin{eqnarray}
p(\yy|A, c, w, \kappa) &=& \prod_{i=1}^N
\frac{1}{(\kappa\sigma)\sqrt{2\pi}}
\exp
\left[
-\frac{1}{2(\kappa\sigma)^2}\left(y_i - \mathcal{M}(x_i; A, c, w)\right)^2
\right] \\
&\propto& \kappa^{-N}\exp\left[-\frac{1}{2(\kappa\sigma)^2}\sum_{i=1}^N\left(y_i - \mathcal{M}(x_i; A, c, w)\right)^2\right]
\end{eqnarray}
We ran this model through DNest and found that it also does not work. The
true value of $F$ is 1, and the inferred values of $F$ (posterior mean
$\pm$ posterior standard deviation) are
$(0.9407 \pm 0.0044)$ and $(0.9402 \pm 0.0077)$
respectively. The true value is a $13.5$-$\sigma$ outlier from the posterior
using the simple model, and this is only reduced to $7.7$-$\sigma$ by the
inclusion of the ``sigma-boost'' parameter $\kappa$. Clearly this is an
unsatisfactory situation.

@@

>>>>

\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\parindent=0cm
\newcommand{\yy}{\mathbf{y}}
\newcommand{\dnest}{\texttt{DNest}}

@@

\begin{abstract}
We discuss the practical use of simply parameterised models in Bayesian data analysis.
These models can be problematic when they are fit to large, informative data sets, resulting in unrealistically small uncertainties on the resulting inferences. In common parlance, the formal statistical errors may be small, but the systematic errors are often large and in many cases, difficult to quantify at all. In principle, the best solution to this problem is to develop more realistic complex models that accurately describe our prior beliefs. However, this is often infeasible in practice, due to computational limitations and the difficulty of defining satisfactory prior distributions over high dimensional spaces. As a pragmatic alternative, we show that it is possible to reinterpret the use of simple models in a way that avoids the problem of overconfident
inferences. The key idea is to weaken the connection between the parameters and the data that is assumed in the prior probabilities. This weakening can be implemented straightforwardly after-the-fact using the Nested Sampling algorithm.
\end{abstract}

@@

When the number of parameters being inferred is 3 or greater, enumeration of
all possible values and their probabilities becomes computationally prohibitive.
In this situation,
Markov Chain Monte Carlo (MCMC) methods are commonly used to produce samples
drawn from the posterior distribution. These samples can then be used to view
the marginal distribution for any particular parameter(s) of interest.
In addition, if the
samples are viewed as a movie, they can give the viewer an intuitive sense of
the uncertainty in the determination of the parameters: any feature of the
models that varies between movie frames is a conclusion that is uncertain, and any
feature that is common to all frames in the movie is a feature that has a high
posterior probability.

@@

\section{An Example: Fitting an Emission Line}
Throughout this paper we will use a simple example of fitting an emission line
to noisy data to explore the properties of simply parameterised models. The
simulated data was generated from the following complex model:
\begin{eqnarray}
\mathcal{M}_0(x) \propto 3\exp\left[-\frac{1}{2}\left(\frac{x}{0.3}\right)^2\right]
+ \exp\left[-\frac{1}{2}\left(\frac{x - 1}{1}\right)^2\right]
+ 5\exp\left[-\frac{1}{2}\left(\frac{x + 1}{0.5}\right)^2\right]
\label{eqn:truth}
\end{eqnarray}
The fluxes were evaluated at $x$-values ranging from $-10$ to $10$ in steps
of $0.001$, and then normalised to sum to 1:
\begin{eqnarray}
\sum_{i=1}^N \mathcal{M}_0(x_i) &=& 1.
\end{eqnarray}
To simulate noise in the measurement process, errors $\{\epsilon_i\}$ were
added, generated
independently from a normal distribution with mean 0 and standard deviation
$\sigma = 3 \times 10^{-4}$.
This produces noisy data $\mathbf{y} = \{y_1, y_2, ..., y_N\}$:
\begin{eqnarray}
\begin{array}{lccr}
y_i = \mathcal{M}_0(x_i) + \epsilon_i, & & & \epsilon_i \sim \mathcal{N}(0, \sigma^2).
\end{array}
\end{eqnarray}
The simulated noisy data is shown in Figure~\ref{emission_line}.

@@

\subsection{The Failure of the Simply Parameterised Model}
We ran the simply parameterised model on the simulated data using the Diffusive
Nested Sampling algorithm \citep{dnest}
\footnote{A C++ implementation of Diffusive Nested Sampling is available
online at {\tt http://github.com/eggplantbren/DNest3/}.}. This provides samples
from the posterior distribution for the parameters $A$, $c$ and $w$. Since these
parameters do not ``exist'' in reality (i.e. the data were generated from the
complex model), it does not make much sense to look at the posterior distribution
for them. However, consider the total flux:
\begin{eqnarray}
F = \sum_{i=1}^N \mathcal{M}(x_i)
\end{eqnarray}
This can be computed both for the true underlying complex model, and the fitted
simple models. In simpler terms, it ``exists''. We computed the posterior distribution
for the total flux using the posterior samples. These are plotted in
Figure~\ref{failure}.

A commonly suggested technique \citep[e.g.][]{gregory} to avoid this situation
is to add an ``error-bar boost'' parameter which we will call $\kappa$.
This adds one more parameter to the list, and the likelihood becomes:
\begin{eqnarray}
p(\yy|A, c, w, \kappa) &=& \prod_{i=1}^N
\frac{1}{(\kappa\sigma)\sqrt{2\pi}}
\exp
\left[
-\frac{1}{2(\kappa\sigma)^2}\left(y_i - \mathcal{M}(x_i; A, c, w)\right)^2
\right] \\
&\propto& \kappa^{-N}\exp\left[-\frac{1}{2(\kappa\sigma)^2}\sum_{i=1}^N\left(y_i - \mathcal{M}(x_i; A, c, w)\right)^2\right]
\end{eqnarray}
We ran this model through \dnest and found that it also does not work. The
true value of $F$ is 1, and the inferred values of $F$ (posterior mean
$\pm$ posterior standard deviation) are
$(0.9407 \pm 0.0044)$ and $(0.9402 \pm 0.0077)$
respectively. The true value is a $13.5$-$\sigma$ outlier from the posterior
using the simple model, and this is only reduced to $7.7$-$\sigma$ by the
inclusion of the ``sigma-boost'' parameter $\kappa$. Clearly this is an
unsatisfactory situation.

@@

\section{Simply-Parameterised Model Parameters Exist}
Consider our emission line example. The true underlying function was given
in Equation~\ref{eqn:truth}. Now imagine that we had a family of simply
parameterised models $M(x)$, parameterised by $\theta$:
\begin{eqnarray}
M(x; \theta)
\end{eqnarray}
Suppose we wanted to, for summary purposes, describe the complex truth
$\mathcal{M}_0(x)$ by some fitted simple model that maximises the following:
\begin{eqnarray}
S(\theta) &=& \int \mathcal{M}_0(x)\log\left(M(x; \theta)\right) \, dx
\end{eqnarray}
subject to the constraint of equal normalisations:
\begin{eqnarray}
\int M(x) \, dx = \int \mathcal{M}_0(x) \, dx
\end{eqnarray}
